---
title: "R Workshop"
author: "Vanessa"
date: '2019-01-21'
output:
  html_document: default
  pdf_document: default
---

## Creating Objects in R

```{r setup, warning=FALSE}

```


Super basic, using R as a fancy calculator:

```{r arithmetic}

3+5

12/7

5*5

# assign value to object/variable
weight_kg <- 55

2.2*weight_kg

weight_lb <- 2.2*weight_kg

sqrt(weight_kg)

round(pi)

round(3.14159)

round(3.14159, digits=2)

round(pi, digits = 6)

round(pi, 10)
```

### Vectors and Data Types

This section will describe some basic data types in R:

```{r}
weight_g <- c(50, 60, 65, 82)

animals <- c("mouse", "rat", "dog")

```

Vector types in R:

* numeric
* character
* logical (TRUE or FALSE)
* factors (categorical data i.e. species)
* Dates

A vector is a data structure in R.

Other data structures:

* lists
* data frames
* matrices
* of course vectors

Often you want to convert lists and matrices to data frames or vectors.

## Data Frames

Next we're going to look at the structure of Data Frames.

```{r}
library(tidyverse)

download.file(url="https://ndownloader.figshare.com/files/2292169", destfile = "read_data/portal_data_joined.csv")

library(here)
#this package makes working directories and file paths easy

surveys <- read_csv(here("read_data", "portal_data_joined.csv"))
# THIS PART IS NEW AND RELEVANT TO ME 

str(surveys)

dim(surveys)
nrow(surveys)
ncol(surveys)

summary(surveys)
```

### Indexing and Subsetting Data Frames

First let's use square bracket subsetting.

Square brackets are great for defining coordinates to extract data from. But what happens when the structure of the data frame changes.

```{r}

#first define the row coordinate and then the column
#also write row and then column
surveys[1, 1]

surveys[1, 6]

#defining only which element we want will return a data frame
surveys[1]

surveys[1:3, 7]

#give us all the rows and columns except column 7
surveys[, -7]

surveys[, -c(1:3)]
```

## Data Manipulation

Key functions for data manipulation:

* `select()`: subsetting columns
* `filter()`: subsets of rows based on conditions
* `mutate()`: create new columns, based on information from other columns
* `group_by()`: creates groups based on categorical data in a column
* `summarize()`: creates summary stats on grouped data
* `arrange()`: sort results
* `count()`: gives a count of discrete values

```{r}
select(surveys, plot_id, species_id, weight)

#negative subsetting
select(surveys, -record_id)

filter(surveys, year==1995,
       species_id=="NL")
```

## Pipes

Pipes allow you to chain together dplyr functions.

Pipe: %>% or cmd-shift-m

```{r}
#write multiple arguments in a sentence using pipes
surveys %>%
  filter(weight<5) %>%
  select(species_id, sex, weight)

surveys_sml <- surveys %>%
  filter(weight<5) %>%
  select(species_id, sex, weight)
```

Challenge #1

Using pipe, subset the surveys dataframe to include animals collected 1995 and retain only the columns year, sex and weight.

```{r challenge 1}
surveys %>%
  filter(year==1995) %>%
  select(year, sex, weight)
```

```{r}
surveys %>%
  mutate(weight_kg=weight/1000,
         weight_kg2=weight_kg*2)

surveys <- surveys %>%
  drop_na(weight) %>%
  mutate(mean_weight=mean(weight))
```

Challenge #2

Using the surveys data from create a new data frame that contains only the species_id column, has a new column called hindfoot_half: contains values that are half the hindfoot_length values. Also, in the new hindfoot_half column there are no NAs and values are all less than 30.

```{r}
surveys_hindfoot_half <- surveys %>%
  drop_na(hindfoot_length) %>% 
  mutate(hindfoot_half=hindfoot_length/2) %>%
  filter(hindfoot_half<30) %>% 
  select(species_id, hindfoot_half, hindfoot_length)
```

```{r}
surveys %>%
  group_by(sex) %>%
  summarize(mean_weight=mean(weight, na.rm = TRUE))

surveys %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight=mean(weight, na.rm=TRUE),
            min_weight=min(weight, na.rm=TRUE)) %>%
  arrange(desc(min_weight))
```

```{r}
surveys %>%
  count(sex, sort = TRUE)

#the above code is synonomous with
surveys %>% 
  group_by(sex) %>%
  summarise(count=n())
```

Challenge # 3

How many animals were caught in each plot_type surveyed.

```{r}
surveys %>%
  count(plot_type)
```


Use group_by and summarize to find the mean, min and max of hindfoot
length (using species_id) for each species. Also, add the number of
observations (hint: see ?n)

```{r}
surveys %>%
  group_by(species_id) %>%
  summarise(mean_length=mean(hindfoot_length, na.rm = TRUE),
            min_length=min(hindfoot_length, na.rm = TRUE),
            max_length=max(hindfoot_length, na.rm = TRUE), n=n())
```

What was the heaviest animal measured in each year? Return the columns
year, genus, species_id and weight.

```{r}
surveys %>%
  group_by(year) %>%
  summarise(genus=first(genus),
            species_id=first(species_id),
    max_weight=max(weight, na.rm=TRUE)) %>% 
  select(year, genus, species_id, max_weight)
#my version
#incorrect! grabbed the first name and filled in the relevant value
#all neotoma except 1977 where it displayed dipodomys because no neot.

max_weights <- surveys %>%
  drop_na(weight) %>%
  group_by(year) %>%
  filter(weight==max(weight)) %>%
  select(year, genus, species_id, weight) %>%
  arrange(year) %>%
  unique()
#brett's version - correct way to do it! wouldn't thought of filter
```

## Export Our Data

```{r}
write_csv(max_weights, here("write_data", "max_weights.csv"))
```

Git history is stored locally here. Hit diff/history for more details.

# Day 2

## Tidy Data in Spreadsheets

Today we are going to look at tidying data.

The functions we use for tidying data are:

* tidyr::spread()
* tidyr::gather()

Note: order of libraries being loaded matters for select, summarize, etc.

### spread()

spread() takes three principle arguments:

1. the data
2. the _key_ column variable will become the new column names
3. the value column variable which will fill the new column variables

We're going to use the surveys dataset

```{r}
library(tidyverse)
library(here)

surveys <- read_csv(here("read_data", "surveys.csv"))

#create a wide data format of surveys using spread

surveys_gw <- surveys %>%
  drop_na(weight) %>%
  group_by(species_id) %>%
  summarise(mean_weight=mean(weight))

str(surveys_gw)

#now we want to spread the dataset out into a wide format

wide_surveys_gw <- surveys_gw %>%
  spread(key=species_id, value=mean_weight)
#long format is usually better, group together species for analysis

#now we are going back to long from the wide dataset that we just created
```

### gather()

gather() takes four arugments:

1. data
2. key
3. value
4. names of columns we use to fill the key variable (or drop)

```{r}
long_surveys <- wide_surveys_gw %>%
  gather(key=species_id, value=mean_weight)
```

## Sending Tidy Data

### Changelog

* Update your changelog with changes you make to your raw data or other significant changes or additions to your projects!

### Data Dictionary

* Create a data dictionary to define our variables

```{r}
tidygsi <- read_csv(here("write_data", "tidy_gsi.csv"))
#noticed an error because comment in column wasn't deleted and created NA's
```

To sync with GitHub: commit then pull then push (not pull then commit then push?)

Having trouble connecting github and local computer when project created locally...

But I think it's fixed now. Create Github project then upload local files there.